---
title: 'Homework 04: Tidy data and joins'
author: "Roger Yu-Hsiang Lo"
date: '2018-10-09'
output:
  github_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = 'center')
```

## Bring rectangular data in

- Load the `Gapminder`, `tidyverse`, and `GGally`:
```{r, warning = FALSE}
#install.packages("GGally")
library(gapminder)
library(tidyverse)
library(GGally)
```

- Some sanity check to make sure the `Gapminder` data was loaded properly:
```{r}
head(gapminder) %>%
  knitr::kable(.)
```

## Data reshaping

The life expectancy of three Asian countries (i.e., Japan, South Korea, and Taiwan) and three African countries (i.e., Benin, Ghana, and Togo) over the years is tabulated below:

```{r}
gapminder %>%
  filter(country %in% c("Japan", "Korea, Rep.", "Taiwan", "Benin", "Ghana", "Togo")) %>%
  select(country, year, lifeExp) %>%
  spread(key = country, value = lifeExp) %>%
  select(year, "Japan", "Korea, Rep.", "Taiwan", "Benin", "Ghana", "Togo") %>%  # Reorder the columns
  knitr::kable()
```

We can see the correlation of life expectancy between different countries by using a correlation plot:

```{r}
gapminder %>%
  filter(country %in% c("Japan", "Korea, Rep.", "Taiwan", "Benin", "Ghana", "Togo")) %>%
  select(country, year, lifeExp) %>%
  spread(key = country, value = lifeExp) %>%
  select(year, "Japan", "Korea, Rep.", "Taiwan", "Benin", "Ghana", "Togo") %>%
  ggpairs(data = ., columns = 2:7)
```

As shown in the plot, life expectancy over the years is highly correlated across different countries, even countries in different continents.

## Data join













