---
title: 'Homework 04: Tidy data and joins'
author: "Roger Yu-Hsiang Lo"
date: '2018-10-09'
output:
  github_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.align = 'center')
```

## Bring rectangular data in

- Load `Gapminder` and `tidyverse`:
```{r}
library(gapminder)
library(tidyverse)
```

- Some sanity check to make sure the `Gapminder` data was loaded properly:
```{r}
head(gapminder) %>%
  knitr::kable()
```

## Data reshaping

### Life expectancy from a few countries over the years

The life expectancy of three Asian countries (i.e., Japan, South Korea, and Taiwan) and three African countries (i.e., Benin, Ghana, and Togo) over the years is tabulated below:

```{r}
gapminder %>%
  filter(country %in% c("Japan", "Korea, Rep.", "Taiwan", "Benin", "Ghana", "Togo")) %>%
  select(country, year, lifeExp) %>%
  spread(key = country, value = lifeExp) %>%
  select(year, "Japan", "Korea, Rep.", "Taiwan", "Benin", "Ghana", "Togo") %>%  # Reorder the columns
  knitr::kable()
```

We can see the correlation of life expectancy between different countries by using a correlation plot from `GGally`:

```{r}
#install.packages("GGally")
library(GGally)

gapminder %>%
  filter(country %in% c("Japan", "Korea, Rep.", "Taiwan", "Benin", "Ghana", "Togo")) %>%
  select(country, year, lifeExp) %>%
  spread(key = country, value = lifeExp) %>%
  select(year, "Japan", "Korea, Rep.", "Taiwan", "Benin", "Ghana", "Togo") %>%
  ggpairs(columns = 2:7) +
  theme_bw() +
  labs(title = "Correlation plot between countries") +
  theme(plot.title = element_text(hjust = 0.5),
        legend.position = "none")
```

As shown in the plot, life expectancy over the years is highly correlated across different countries, even countries in different continents.

### Life expectancy growth in different continents

Let us compare the five-year growth in life expectancy (since the data were sampled every five years) in different continents.

```{r}
gapminder %>%
  group_by(country) %>%
  mutate(growth_lifeExp = lifeExp - lag(lifeExp)) %>%  # Calculate the growth compared to the previous year
  filter(year != 1952, continent != "Oceania") %>%  # Remove NA values and data from Oceania
  ungroup() %>%
  group_by(year, continent) %>%
  summarize(mdn_grt_lifeExp = median(growth_lifeExp)) %>%  # Calculate the median
  spread(key = continent, value = mdn_grt_lifeExp) %>%
  knitr::kable()
```

We can already see a rough trend from the table: the growth in life expectancy decreased slightly over the years across continents. The trend becomes more "visible" when the data are plotted.

```{r}
gapminder %>%
  group_by(country) %>%
  mutate(growth_lifeExp = lifeExp - lag(lifeExp)) %>%  # Calculate the growth compared to the previous year
  filter(year != 1952, continent != "Oceania") %>%  # Remove NA values and data from Oceania
  ungroup() %>%
  group_by(year, continent) %>%
  ggplot(aes(x = year, y = growth_lifeExp, color = continent)) +
  facet_wrap(~continent) + 
  geom_jitter(alpha = 0.1) +
  geom_smooth(method = "lm") +  # Fit regression lines
  scale_x_continuous(breaks = seq(1957, 2007, 10)) +
  coord_cartesian(ylim=c(-2, 4)) +
  labs(title = "Life expectancy change over years", x = 'Year', y = 'Life expectancy growth (years)') +
  theme(plot.title = element_text(hjust = 0.5),
        legend.position = "none")
```

As shown in the plot, the growth in life expectancy decreased gradually across the four continents. We can also see some differences between continents. Around 1960, the growth is biggest in Asia and smallest in Europe. In addition, there is quite some variation in Africa, compared with Europe or Americas in general, as indicated by the width of the gray ribbons.

## Data join

### Life expectancy and population in 50 years

Let us try to show the data on a map. To do this, we need to first import the package `ggmap`. The data (`country_geo_info.txt`) of the longitude and latitude of different countries are obtained from [here](https://opendata.socrata.com/dataset/Country-List-ISO-3166-Codes-Latitude-Longitude/mnkm-8ram). The ISO code of each country can be found in the data frame `country_codes` that is loaded along with `Gapminder`.

In what follows, the information about the population and life expectancy of each country in 1957 and 2007 is plotted on a world map.

```{r, fig.width = 12, fig.height = 4}
library(ggmap)

# Read in the data that contain longitude and latitude information
country_info <- read.csv("country_geo_info.txt", header = TRUE, sep = "\t")

# Remove variables that we do not need
country_info <- country_info %>%
  select(iso_num, latitude, longitude)

# Extract variables from country_codes that we need
country_iso_num <- country_codes %>%
  select(country, iso_num)

gapminder %>%
  filter(year %in% c(1957, 2007)) %>%  # Get entries from 1957 and 2007
  left_join(., country_iso_num, by = "country") %>%  # Add a column with iso code for each country
  left_join(., country_info, by = "iso_num") %>%  # Add two columns: latitude, longitude
  ggplot(aes(x = longitude, y = latitude, size = pop, color = lifeExp)) +
  theme_bw() +  # Change theme
  facet_wrap(~ year) +
  borders("world", colour = "gray80", fill = "gray75") +  # Add a gray world map in the background
  geom_point() +  # The size of points represents pop, and the color the life expectancy
  scale_size_area() +  # Make the size of points corresponds to pop
  scale_size_continuous(name = "Population") +
  scale_color_continuous(name = "Life exp.") +
  labs(title = "Population and life expectancy in 1957 and 2007") +
  theme(plot.title = element_text(hjust = 0.5),  # Center the title
        axis.title = element_blank(),  # Remove axis labels and ticks
        axis.ticks = element_blank(),
        axis.text = element_blank())
```

Comparing the two maps, it is easy to see that both population and life expectancy increased over time in each continent, with a significant improvement of life expectancy in Asia and South America.
